---
title: 'ATAC_04 - Calling nucleosome-free regions with MACS2'
author: "Emma Bell"
date: '2020-04-30'
output:
  html_document:
    df_print: paged
    toc: yes
    toc_float: true
    toc_depth: 2
    fig_caption: true
---

# Summary

* Aim: To call nucleosome-free regions with MACS2.
* Objectives:
  1. To use MACS2 to call nucleosome-free regions.

## Configuration

This analysis requires the following packages:

```{r}
library(readxl)
library(vtree)
library(writexl)
```

This analysis requires the following R configuration object:

```{r}
r.config <- readRDS(file = file.path("code","r_config.RDS"))
```

## Metadata

```{r}
metadata <- read_excel(path = file.path(r.config$dirs$data.dir, "metadata.xlsx"), sheet = "ATAC")
metadata
```

I'll use the `vtree` package to visualise the experimental design.

```{r}
vtree(z = metadata, vars = c("Stimulated","DAC"))
```

# Analysis

## 1. Using MACS2 to call nucleosome-free regions

These are the arguments to MACS2 we used to call nucleosome-free regions.

```{bash eval=FALSE}
macs2 callpeak -t $BAM -g hs -n $OUTNAME -q 0.25 --nomodel --shift -100 --extsize 200
```

The resulting narrowPeak files are located at `data/03_augmented/03_atac_nfrs/01_narrowPeaks`.
